---
title: "Assignment 2 final"
author: "King Banaian"
date: "2025-02-21"
output:
  word_document: default
  pdf_document: default
---
##Loading needed libraries

```{r setup, include=FALSE}
library(ggplot2)
library(forecast)
library(tseries)
library(knitr)
library(lubridate)
knitr::opts_chunk$set(echo = TRUE)
```

## Loading the data

The method I used above includes the forecast library, which allows me to use some slightly different ways of creating the training data and the testind datasets. 

```{r read data and convert to time series}
# Read and prepare the data
data <- read.csv("USTRADE.csv")
data$observation_date <- as.Date(paste0(data$observation_date, "-01"), format="%m/%d/%Y-%d")

# Convert to time series object
ts_data <- ts(data$USTRADE, frequency=12, 
              start=c(year(min(data$observation_date)), month(min(data$observation_date))))

summary(ts_data) 
```

## Splitting the data

Now we will create the training data through the end of 2021, set up the testing period of 2022-23, and then a last set to hold 2024:

```{r data windows}
# Split into training (up to Dec 2021) and test (2022-2023)
train_data <- window(ts_data, end=c(2021,12))
test_data <- window(ts_data, start=c(2022,1), end=c(2023,12))
```

## Forecasts for 2022-23

We now generate the forecasts and use the accuracy() function to collect forecast summary statistics.

```{r}
# Generate forecasts
naive_fc <- naive(train_data, h=24)
snaive_fc <- snaive(train_data, h=24)
drift_fc <- rwf(train_data, h=24, drift=TRUE)

# Calculate accuracy measures
naive_accuracy <- accuracy(naive_fc, test_data)
snaive_accuracy <- accuracy(snaive_fc, test_data)
drift_accuracy <- accuracy(drift_fc, test_data)
```

##Create a pretty table to compare the forecast error statistics for 2022-2023

I'm going to make a data frame here report out three important error statsitics: the root mean squared error (RMSE), the mean absolute error (MAE) and the mean absolute percent error (MAPE). I wasn't very fussy in how I graded what you did to make this table. I just wanted to see how you would present the results.

```{r forecast table}
# Calculate accuracy measures
naive_accuracy <- accuracy(naive_fc, test_data)
snaive_accuracy <- accuracy(snaive_fc, test_data)
drift_accuracy <- accuracy(drift_fc, test_data)

# Create comparison table
comparison_table <- data.frame(
  Method = c("Naive", "Seasonal Naive", "Drift"),
  RMSE = round(c(naive_accuracy[2,"RMSE"], 
                 snaive_accuracy[2,"RMSE"],
                 drift_accuracy[2,"RMSE"]), 2),
  MAE = round(c(naive_accuracy[2,"MAE"],
                snaive_accuracy[2,"MAE"],
                drift_accuracy[2,"MAE"]), 2),
  MAPE = round(c(naive_accuracy[2,"MAPE"],
                 snaive_accuracy[2,"MAPE"],
                 drift_accuracy[2,"MAPE"]), 2)
)

# Print table
print("Forecast Accuracy Comparison:")
print(kable(comparison_table))
```

As you can see, the drift model is better across all three forecast accuracy summary statistics. We will use this to create our final forecast for 2024.

##Make 2024 forecast

In this section I have added a graph that shows how 2024 actually turned out (since I gave you those data) to your forecast.

```{r 2024 forecast, echo=FALSE}
# Generate 2024 forecasts using best method (based on RMSE) using data through 2023
data_for_forecast <- window(ts_data, end=c(2023,12))
final_forecast <- rwf(data_for_forecast, h=12, drift=TRUE)
print("\n2024 Forecasts (Drift Method):")
print(forecast(final_forecast))

# Plot the forecast with actual 2024 data
autoplot(final_forecast) +
  autolayer(window(ts_data, start=c(2024,1), end=c(2024,12)), series="Actual 2024") +
  theme_minimal() +
  labs(x = "Time", y = "US Retail Trade (Millions of Dollars)", 
       title = "US Retail Trade: 2024 Forecast vs Actual") +
  coord_cartesian(xlim = c(2023, 2024.99),
                 ylim = c(15000, 17000)) +
  scale_x_continuous(breaks = seq(2023, 2024.99, 0.25)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")
```

The error bands are quite large, but the forecast tracks pretty well to the actual data, as you can see on the graph I've drawn. The way I have done the forecast only uses data through 2023 (updating the drift method's parameters from 2021). 